### 2021-10-29 documents/sections/verbs.tex
more work on 'to go down'

\subsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{k-} in \PPek \pcref{sec:pekodian} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
The resistance against the former extension was overcome in \bakairi.
In \akuriyo, the verb already had an irregularly inflected first person form with \obj{p-}, inherited from \PTir.

@@

\pex<gowhere>
\a<car-18> \carijo \parencite[][102]{guerrero2019carijo}\\
\begingl
\gla əji-wa-e j-eh-ɨ//
\glb \gl{2}-search-\gl{sup} \gl{1}-come-\gl{pfv}//
\glft \qu{I came looking for you.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
However, a look at the comparative picture suggests a much more complicated situation.
\cref{tab:godown} shows all attested cognates, some already discussed, including verb class membership and aligned segments.
While a form \rc{ɨpɨtə} can be reconstructed to \PC, different (proto-)languages do not agree about the class of this verb.
The reflexes are fairly evenly split between \gl{s_a_} and \gl{s_p_} -- in those languages that preserve the split-\gl{s} system.

@@

\pex<gowhere>
\a<car-18> \carijo \parencite[][102]{guerrero2019carijo}\\
\begingl
\gla əji-wa-e j-eh-ɨ//
\glb \gl{2}-search-\gl{sup} \gl{1}-come-\gl{pfv}//
\glft \qu{I came looking for you.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}\gl{pro} \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
Its causativized form is \obj{ɨptəka} \parencite[255]{wayanatavares2005}, containing a reflex of the transitivizer \rc{-ka}, which was restricted to \gl{s_p_} verbs in \PC \parencite{gildea2019overview}.
This leads us to posit the hypothesis that the verb was a regular member of the \gl{s_p_} class in pre-\wayana, but partially switched to the \gl{s_a_} class.
This in turn implies that all reflexes of this verb with \gl{s_a_} status were members of the \gl{s_p_} class at some point, but switched classes.

@@

These circumstances make it impossible to answer the question of whether it was affected by the extensions in \PTir, \carijo, and \yukpa.
For \PTir, we cannot establish a relative chronology of the verb class change, the introduction of the idiosyncratic marker \rc{p-}, and the extension of \rc{t-}.
It could have been an \gl{s_p_} verb at the time of the extension, it could have been an \gl{s_a_} verb and remained unaffected, or it could have been affected, but with a subsequent erasure of the results by the introduction of \rc{p-}.
For \carijo and \yukpa, we cannot know whether the verb switched verb class before the breakdown of the entire split-\gl{s} system -- this is a plausible scenario especially in the former case.
In all three cases, it is possible that the verb had \gl{s_a_} status at the time of the extension, but it is also possible that it was not even a potential target due to its \gl{s_p_} status.
On the other hand, the class change must have happened before the split-up of \trio and \akuriyo, and therefore this verb resisted the extension of \akuriyo \obj{k-}.
Likewise, the class change must have taken place before the extension of \PPek \rc{k-}, since other scenarios would require an \gl{s_p_} verb taking on regular \gl{s_a_} markers, but selecting the archaic first person marker.

>>>>

\subsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{k-} in \PPek \pcref{sec:pekodian} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
The resistance against the former extension was subsequently overcome in \bakairi; its fate in \ikpeng is unknown.
When \akuriyo extended \obj{k-}, the verb already had an irregularly inflected first person form with \obj{p-}, inherited from \PTir.

@@

\pex<gowhere>
\a<car-32> \carijo \perscommpar{David Felipe Guerrero}\\
\begingl
\gla irə wat͡ʃinakano tae j-ehɨtə-e//
\glb \gl{inan}.\gl{ana} body.of.water along.bounded \gl{1}-go.down-\gl{npst}//
\glft \qu{…I go down through that guachinacán.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}\gl{pro} \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
However, a look at the comparative picture suggests a much more complicated situation.
\cref{tab:godown} shows all attested forms, including verb class membership where applicable, as well as aligned segments for forms we judge to be fully cognate.
For those forms, \rc{ɨpɨtə} can be reconstructed to \PC, but different (proto-)languages do not agree about the class of this verb.
The reflexes are fairly evenly split between \gl{s_a_} and \gl{s_p_} -- in those languages that preserve the split-\gl{s} system.

@@

\pex<waygodown>\wayana \parencite[][200]{wayanatavares2005}
\a<way-73>
\begingl
\glpreamble ïwïptëë//
\gla ɨ-w-ɨptə-rɨ//
\glb \gl{1}-\gl{s_a_}-go.down-\gl{nmlz}//
\glft \qu{my going down}//
\endgl
\a<way-71>
\begingl
\gla əw-ɨptə-k//
\glb \gl{2}-go.down-\gl{imp}//
\glft \qu{Go down!}//
\endgl
\xe
%
Its causativized form is \obj{ɨptə-ka} \parencite[255]{wayanatavares2005}; the transitivizer \rc{-ka} was restricted to \gl{s_p_} verbs in \PC \parencite{gildea2019overview}.
This leads us to posit the hypothesis that the verb was a regular member of the \gl{s_p_} class in pre-\wayana, but partially switched to the \gl{s_a_} class.
This in turn implies that all reflexes of this verb with \gl{s_a_} status were members of the \gl{s_p_} class at some point, but switched classes.

@@

These circumstances make it impossible to answer the question of whether \qu{to go down} was affected by the extensions in \PTir, \carijo, and \yukpa.
For \PTir, we cannot establish a relative chronology of the verb class change, the introduction of the idiosyncratic marker \rc{p-}, and the extension of \rc{t-}.
For \carijo and \yukpa, we cannot know whether the verb switched class before the breakdown of the entire split-\gl{s} system.
While there is no language-internal evidence for such a switch, \qu{to go down} does have an inclination for class switches; in the case of \carijo, that could have already happened at the \PTar stage.
In all three cases, it is possible that the verb had \gl{s_a_} status at the time of the extension, resisting and keeping its old marker, but it is also possible that it was not even a potential target due to its \gl{s_p_} status.
On the other hand, the class change must have happened before the split-up of \trio and \akuriyo, and therefore this verb resisted the extension of \akuriyo \obj{k-}.
Likewise, it seems very likely that the class change took place before the extension of \PPek \rc{k-}.
Otherwise, the newly-turned-\gl{s_a_} verb would have taken on archaic and lexically heavily restricted \rc{w-}, either in \PTar, \PXin, or \arara.
